---
title: DATA 606 Data Project Proposal
author: 
---

### Data Preparation

```{r setup, echo=TRUE, results='hide', warning=FALSE, message=FALSE}
# Imports
library(dplyr)
library(ggplot2)

# Set up our url
crash_url <- "https://raw.githubusercontent.com/riverar9/cuny-msds/main/data606/project/01_proposal/Crash%20Reporting.csv"

# Read in the url using read.csv
crash_df <- read.csv(
  crash_url
)
# Inspect the raw dataframe
head(crash_df)

# Since the dataframe is large, let's filter it to what we're interested in
crash_df <- crash_df |>
  select(
    Vehicle.Body.Type,
    ACRS.Report.Type,
    Crash.Date.Time
  ) |>
  filter(
    Vehicle.Body.Type != ""
  )

crash_df |>
  group_by(ACRS.Report.Type) |>
  summarise(count = n())

total_report_count <- nrow(crash_df)
```


### Research question 

**You should phrase your research question in a way that matches up with the scope of inference your dataset allows for.**

 Is there a difference in the crash rate between cars of different body types for Fatal and Injury crashes?

### Cases 

**What are the cases, and how many are there?**

Each case represents one vehicle crash incident report. There are `r format(total_report_count, big.mark = ",")` observations in this dataset.

### Data collection 

**Describe the method of data collection.**

This data was retrieved from (https://data.gov/)[https://data.gov/]. This is a resource from the United States Federal Government that provides datasets for free use.

### Type of study 

**What type of study is this (observational/experiment)?**

This is an observaitonal study.

### Data Source 

**If you collected the data, state self-collected. If not, provide a citation/link.**

This data was obtained from Data.gov.

(URL to the data used.)[https://catalog.data.gov/dataset/crash-reporting-drivers-data]

### Dependent Variable

**What is the response variable? Is it quantitative or qualitative?**

Our Dependent variable will be the number of crashes and it is quantitative.

### Independent Variable(s)

Our Independent variable will be: 
* Vehicle.Body.Type.


### Relevant summary statistics 

**Provide summary statistics for each the variables. Also include appropriate visualizations related to your research question (e.g. scatter plot, boxplots, etc). This step requires the use of R, hence a code chunk is provided below. Insert more code chunks as needed.**

```{r, fig.width = 10, fig.height = 6}
frequency_data <- crash_df |>
  count(Vehicle.Body.Type)

ggplot(
  frequency_data,
  aes(
    x = n,
    y = reorder(
      Vehicle.Body.Type,
      n
    )
  )
) +
  geom_bar(
    stat = "identity",
    fill = "steelblue"
  ) +
  labs(
    title = "Car Crash Frequency by Vehicle Type",
    x = "Count of Incidents",
    y = "Vehicle Type"
  )

vehicle_summary <- crash_df |>
  group_by(
    Vehicle.Body.Type
  ) |>
  summarise(
    vehicle_occurances = n()
  )

vehicle_crash_summary <- crash_df |>
  mutate(
    injury_fatal_crash = ifelse(
      ACRS.Report.Type == "Fatal Crash"
      | ACRS.Report.Type == "Injury Crash",
      "Fatal/Injury",
      "Property"
    )
  ) |>
  group_by(
    Vehicle.Body.Type,
    injury_fatal_crash
  ) |>
  summarise(
    vehicle_crash_occurances = n()
  )

vehicle_crash_summary <- left_join(
  vehicle_crash_summary,
  vehicle_summary,
  by = "Vehicle.Body.Type"
)

vehicle_crash_summary <- vehicle_crash_summary |>
  mutate(
    crash_freq = vehicle_crash_occurances / vehicle_occurances
  )

ggplot(
  data = vehicle_crash_summary,
  aes(
    x = crash_freq,
    y = reorder(
      Vehicle.Body.Type,
      crash_freq
    ),
    fill = injury_fatal_crash
  )
) +
  geom_bar(
    stat = 'identity'
  ) +
  labs(
    x = "Frequency",
    y = "Vehicle Body Type",
    title = "Proportion of Fata/Injury Accidents"
  )
```
