---
title: "DATA 624 - Homework 1"
author: "Richie Rivera"
output:
  pdf_document: default
---

# Question 2.1
Explore the following four time series: Bricks from aus_production, Lynx from pelt, Close from gafa_stock, Demand from vic_elec.

```{r library, echo=FALSE, message=FALSE, warning=FALSE}
# Installing the package
#install.packages("fpp3", dependencies = TRUE)

# Importing the library
library("fpp3")

```
```{r loading data}
# Loading the datasets
data(aus_production)
data(pelt)
data(gafa_stock)
data(vic_elec)
```

#### A: Use ? (or help()) to find out about the data in each series.


```{r aus_production, eval=FALSE}
?aus_production
```

```{r pelt, eval=FALSE}
?pelt
```

```{r gafa_stock, eval=FALSE}
?gafa_stock
```

```{r vic_elec, eval=FALSE}
?vic_elec
```

#### B: What is the time interval of each series?

| Series            | Time Interval | Description |
|----------         |----------     | ---------- |
| aus_production    | Quarterly     | Quarterly production of selected commodities in Australia. |
| pelt              | Annual        | Pelt trading records |
| gafa_stock        | Daily         | GAFA stock prices |
| vic_elec          | Half-Hourly   | Half-hourly electricity demand for Victoria, Australia |


#### C: Use autoplot() to produce a time plot of each series.

```{r question1c, fig.width=10, fig.height=5}
autoplot(aus_production, Bricks) + ggtitle(
  "Quarterly Brick production of selected commodities in Australia."
) + geom_line(color = "blue", size = 1.5) +
  theme(axis.text = element_text(size = 12))

autoplot(pelt, Lynx) + ggtitle(
  "Lynx Pelt trading records"
) + geom_line(color = "red", size = 1.5) +
  theme(axis.text = element_text(size = 12))

autoplot(gafa_stock, Close) + ggtitle(
  "GAFA stock Close prices"
) + theme(axis.text = element_text(size = 12))

autoplot(vic_elec, Demand) + ggtitle(
  "Half-hourly electricity demand for Victoria, Australia"
) + geom_line(color = "orange", size = 0.5) +
  theme(axis.text = element_text(size = 12))
```

#### D: For the last plot, modify the axis labels and title.

```{r question1d, fig.width=10, fig.height=5}
autoplot(vic_elec, Demand) + ggtitle(
  "Half-hourly electricity demand for Victoria, Australia"
) + geom_line(color = "green", size = 0.5) +
  theme(axis.text = element_text(size = 12), aspect.ratio = 0.5) +
  xlab("Half-Hour Interval") +
  ylab("Demand [MWh]")
```

# Question 2.2
Use filter() to find what days corresponded to the peak closing price for each of the four stocks in gafa_stock.

```{r question2}
# Importing dplyr
library(dplyr)

head(gafa_stock)

gafa_stock |>
  select(
    Symbol,
    Date,
    Close
  ) |>
  group_by(Symbol) |>
  filter(
    Close == max(Close)
  )
```

# Question 2.3
Download the file tute1.csv from the book website, open it in Excel (or some other spreadsheet application), and review its contents. You should find four columns of information. Columns B through D each contain a quarterly series, labelled Sales, AdBudget and GDP. Sales contains the quarterly sales for a small company over the period 1981-2005. AdBudget is the advertising budget and GDP is the gross domestic product. All series have been adjusted for inflation.

#### A. You can read the data into R with the following script
```{r question2.3a}
# importing readr
library(readr)

# Reading and viewing the csv
tute1 <- read_csv("https://raw.githubusercontent.com/riverar9/cuny-msds/main/data624-predictive-analytics/homework/homework-1/tute1.csv")
View(tute1)
```

#### B. Convert the data to time series
```{r question2.3b}
# Converting the data into a timeseries
mytimeseries <- tute1 |>
  mutate(Quarter = yearquarter(Quarter)) |>
  as_tsibble(index = Quarter)

head(mytimeseries)
```

#### C. Construct time series plots of each of the three series

```{r question2.3c}
# Create a series of plots using facet_grid
mytimeseries |>
  pivot_longer(-Quarter) |>
  ggplot(aes(x = Quarter, y = value, colour = name)) +
  geom_line() +
  facet_grid(name ~ ., scales = "free_y")
```

Check what happens when you don’t include facet_grid()

```{r question2.3c2}
mytimeseries |>
  pivot_longer(-Quarter) |>
  ggplot(aes(x = Quarter, y = value, colour = name)) +
  geom_line()
```

Without facet_grid, the plots are all on the same chart. In my opinion this is more helpful as it provides immediate insight into the relative value of these timeseries against eachother.

# Question 2.4
The USgas package contains data on the demand for natural gas in the US.

#### A. Install the USgas package.

#### B. Create a tsibble from us_total with year as the index and state as the key.

#### C. Plot the annual natural gas consumption by state for the New England area (comprising the states of Maine, Vermont, New Hampshire, Massachusetts, Connecticut and Rhode Island).

# Question 2.5

#### A. Download tourism.xlsx from the book website and read it into R using readxl::read_excel().

#### B. Create a tsibble which is identical to the tourism tsibble from the tsibble package.

#### C. Find what combination of Region and Purpose had the maximum number of overnight trips on average.

#### D. Create a new tsibble which combines the Purposes and Regions, and just has total trips by State.

# Question 2.8
Use the following graphics functions: autoplot(), gg_season(), gg_subseries(), gg_lag(), ACF() and explore features from the following time series: “Total Private” Employed from us_employment, Bricks from aus_production, Hare from pelt, “H02” Cost from PBS, and Barrels from us_gasoline.

#### A. Can you spot any seasonality, cyclicity and trend?

#### B. What do you learn about the series?

#### C. What can you say about the seasonal patterns?

#### D. Can you identify any unusual years?