---
title: "Data 607 - Assignment 1"
author: "Richie R."
output:
  pdf_document: default
  html_document: default
---

# Overview
We will be looking into a dataset I personally like, the tripdata dataset published by citibike of their trips. Specifically, we will look at the data corresponding with December 2023 and look at the most popular start and end location for electric bikes (e-bikes).
This data is not natively available online as a delimited file, so we will need to follow the below instructions to be able to access this dataset as a dataframe:

This dataset can be found at the url below:
    https://s3.amazonaws.com/tripdata/index.html

We will be using the '202312-citibike-tripdata.csv.zip' dataset.

1. Download the zipped file to the working directory
2. Unzip the downloaded dataset 
3. Read the unzipped dataset as a dataframe
4. Create a filtered_df dataset which is a dataset to only include e-bike rides
5. Group filtered_df by the starting location and count the number of distinct Ride IDs and sort the data by count descending
6. Repeat step 4, but use the ending location rather than the starting location

## 1. Downloading the dataset to the working directory
```{r Downloading the dataset to the working directory}
file_name       <- "202312-citibike-tripdata.csv"

zip_file_name   <- paste(file_name, ".zip", sep="")

download_url    <- paste("https://s3.amazonaws.com/tripdata/", zip_file_name, sep="")

download.file(download_url, dest=zip_file_name, mode="wb")
```

## 2. Unzipping the downloaded dataset into the same directory

```{r Unzipping the downloaded dataset into the same directory}
unzipped_folder = "data_citibike_rides"
unzip(zip_file_name, exdir=unzipped_folder)
```

## 3. Read in the data as an R dataframe
We'll do so by creating the csv path with the variables established above and then display the first few rows using the head() function.
```{r Read in the data as an R dataframe}
csv_path = paste(unzipped_folder, "/", file_name, sep="")

citi_df = read.csv(csv_path)

head(citi_df)
```

## 4. Create an R dataframe with the subset of the rows which are e-bikes

To do so, we will begin by filtering the dataframe to entries where ridable_type = "electric_bike"

Once that's done, we will only keep the following columns:

- start_station_name

- end_station_name

```{r Create an R dataframe with the subset of the rows which are e-bikes}
filtered_df <- subset(
    citi_df
    , rideable_type == "electric_bike"
    , select = c(start_station_name, end_station_name)
)
```

## 5. With filtered_df, we will look at the most common starting station
To do so, we'll summarize the data grouping by start_station_name and counting the number of times each station shows up.

```{r With filtered_df, we will look at the most common starting station}
library(dplyr)
 
filtered_df %>%
  group_by(start_station_name) %>%
  summarize(Count=n()) %>%
  arrange(desc(Count))

```

From the block above, we can see that '7 Ave & Central Park South' was the most popular starting location in December.

## 6. With fitlered_df, we will look at the most common ending location

```{r With fitlered_df, we will look at the most common ending location}
filtered_df %>%
  group_by(end_station_name) %>%
  summarize(Count=n()) %>%
  arrange(desc(Count))
```

Interestingly enough, the same station ('7 Ave & Central Park South') was also the most popular ending location!

# Conclusion

It appears that '7 Ave & Central Park South' is the most popular start and end station for rides in December 2023.