---
title: "Data 607 - Assignment"
author: "Richie R."
output:
  pdf_document: default
  html_document: default
---

# Overview
In this assignment, we will be tidying a dataset containing delay and ontime information for two airlines by city.

The dataset resembles the table below:

|| |Los Angeles|Phoenix|San Diego|San Francisco|Seattle|
|------|-------|-----------|-------|---------|-------------|-------|
|ALASKA|on time|497        |221    |212      |503          |1,841  |
|      |delayed|62         |12     |20       |102          |305    |
|      |       |           |       |         |             |       |
|AM WEST|on time|694        |4,840  |383      |320          |201    |
|      |delayed|117        |415    |65       |129          |61     |


## 1. Setting up the data

Firstly, we will need to have this data available from somewhere in the format above. I decided to use GCP's BigQuery and initialized the table with the below SQL statements:

```{SQL eval = FALSE}
drop table if exists data607.assignment4_data; # nolint: error.

CREATE TABLE data607.assignment4_data (
row_id int64,
column1 STRING,
column2 STRING,
`Los Angeles` INT64,
`Phoenix` INT64,
`San Diego` INT64,
`San Francisco` INT64,
`Seattle` INT64
);

INSERT INTO data607.assignment4_data 
(row_id, column1, column2, `Los Angeles`, `Phoenix`, `San Diego`, `San Francisco`, `Seattle`) # nolint: line_length_linter.
VALUES
(1,'ALASKA', 'on time',494,221,212,503,1841),
(2,NULL,'delayed',62,12,20,102,305),
(3,NULL,NULL,NULL,NULL,NULL,NULL,NULL),
(4,'AM West', 'on time',694,4840,383,320,201),
(5,NULL,'delayed',117,415,65,129,61);
```

A row_id was necessary as the order of the rows mattered and there was an issue where rows would be ordered by alphabetical order on the first column, disrupting the order.

## 2. Extracting the data from GCP

Since we are using biguery, we will need the path to our file and some parameters:

```{r setting up BigQuery parameters}
library(bigrquery)
library(dplyr)
library(tidyverse)

path_to_bq_creds <- "data607/gcp_bq_auth/gcp_bq_auth.json"

project_id <- "cuny-msds"
dataset_name <- "data607"
raw_data_tablename <- "assignment4_data"

bigrquery::bq_auth(path = path_to_bq_creds)
```

With the set up complete, we can now interact with the database. Let's do so to get the data from our table and order by row_id:

```{r get flight data}
dataset <- bq_dataset(
  project_id, # nolint
  dataset_name
)

flight_table <- bq_dataset_query(
  dataset, # nolint
  paste("SELECT * FROM ", dataset_name, ".", raw_data_tablename)
)

# As a reminder, we want the initial dataframe to preserve the order from
#   the question's prompt.
flight_data <- bq_table_download(flight_table) |>
arrange(row_id)

flight_data
```

## 3. Tidy the data for analysis

To tidy this data, we'll need to do a few things. Firstly, I would like to define how I prefer this data to be oriented. I would prefer that there are 4 columns:

1. Airline
2. City
3. On Time
4. Delayed

In order to get the data into this format, we will need to perform the following transformations to the dataframe:

1. Remove any fully empty rows
2. Establish the names of the columns
3. Clear out any empty cells
4. Pivot the data from a wide to a long format
5. Pivot the data to a wide format where the flight status becomes a column

To remove the empty row, we will say that the status column must be populated. That means that if there is nothing entered in the status column then we will remove it:

```{r remove the empty row}
flight_data <- flight_data |>
  drop_na(column2)# nolint

flight_data
```

Step 2:

```{r rename columns}
colnames(flight_data) <- c(
  'row_id', # nolint
  'airline',
  'flight_status',
  'Los Angeles',
  'Phoenix',
  'San Diego',
  'San Francisco',
  'Seattle'
)

flight_data
```

For step 3, I'm making the assumption that the airline in row 2 is `ALASKA` and `AM West` in row 5. I'm making this assumption based on looking at Phoenix. Assuming that most airline flights are on time (which sometimes doesn't feel that way), we can eliminate row 5 as being part of `ALASKA` as there are 415 delayed flights in row 5 compared to the 221 on time flights for `ALASKA`. Then, by the process of elimination, we can safely assume that row 5 represents `AM West` delayed flights.

Now that we know this, we can use the `fill()` function to fill the missing entries and then remove the row_id:
```{r forward fille}
flight_data <- flight_data |>
  fill(airline) |> # nolint
  select(!row_id)

flight_data
```

For step 4, we can employ the `pivot_longer()` function:

```{r turn wide into long}
flight_data <- flight_data |>
  pivot_longer( # nolint
    cols = !c(airline, flight_status), # nolint
    names_to = "city",
    values_to = "observations"
)

head(flight_data)
```

Finally, for step 5:

```{r turn long data into wide}
flight_data <- flight_data |>
  pivot_wider( # nolint
    names_from = flight_status, # nolint
    values_from = observations
)

colnames(flight_data) <- c('airline','city','on_time','delayed')

flight_data
```

## 4. Flight Analysis

For this portion, we will compare the performance by city for each airline:

```{r compare delay percentage by airline and city}
flights_summary <- flight_data |>
  mutate(total_flights = on_time + delayed) |> # nolint
  mutate(delay_pct = delayed/total_flights)

ggplot(flights_summary, aes(x = airline, y = delay_pct, fill = airline)) +
  geom_col() + # nolint
  facet_wrap(~city) +
  scale_y_continuous(labels = scales::percent) +
  ylab("Delay Percentage") +
  xlab("")
```

This graph above shows pretty clearly that `AM West` pretty reliably has a higher percentage of delayed flights over `ALASKA`. For both airlines, Pheonix was the airport with the lowest delay percentage.

```{r SFO delay comparison}
flights_summary |>
  filter(city == "San Francisco") # nolint

.287/.169
```

Visually, it seems that San Francisco has the greatest difference in on time percentage where `AM West` is about has a delay rate 69.8% greater than that of `ALASKA`.

Looking at the total number of flights:

```{r total flights}
ggplot(flights_summary, aes(x=airline, y=total_flights, fill=airline)) +
  geom_col() + # nolint
  facet_wrap(~city) +
  geom_label(aes(label = round(100*delay_pct,0)))
```

Now looking at this new figure, we have the delay percentage as the text on the bars and the number of flights at each airport. From here we can see that although Phoenix had the lowest delay percentage, it sees the most flights.

# Conclusion

By cleaning up some relatively unstructured data about flight information, we were able to analyze the data and determine that anyone flying `AM West` out of San Francisco can expect to be delayed.

We can see the delay percentage by city and compeitor. From here, we can see that `AM West` consistently performs worse than `ALASKA`. This is disheartening as they have many more flights than `ALASKA`.